Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic

### 2024-03-26 paper.Rmd
ajustes

---
title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
# params:
# lang: en-GB
author: 
- name: Juan Carlos Castillo
  affiliation: Department of Sociology, Universidad de Chile
  email: juancastillov@uchile.cl
  number: 1, 3 
- name: Julio Iturra
  affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
  number: 2
- name: Kevin Carrasco
  affiliation: Center for Social Conflict and Cohesion Studies
  number: 3
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. To what extent are such educational disparities justified in society? And, what are the main factors driving this kind of justification? This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy. The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.

@@

   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
output:
  bookdown::pdf_document2:
    template: null
    number_sections: true    
    toc: false
    keep_tex: false
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
  bookdown::word_document2:
    number_sections: false
    reference_docx: input/template.docx
    toc: false
  bookdown::html_document2:
    number_sections: false
linkcolor: blue
bibliography:
  - input/bib/merit-edjust.bib
csl: input/bib/apa7-singlespace.csl
editor_options:
  chunk_output_type: console
geometry: margin=1in
header-includes:
  - \usepackage{times}           # Times New Roman
  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
always_allow_html: true # for word output
fontsize: 12pt
papersize: a4
---

@@

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H", fig.align = 'center')
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

@@

The Chilean case is particularly relevant when it comes to the study of inequalities in education. The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016]. The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joikomujicaCuasimercadoEducativoChile2019]. Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities. Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramosarellanoEducacionPromesaIncumplida2022]. Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.

@@

Our second argument is based on the evidence that relates rational interests with justifications and preferences [@hing_failure_2019], expecting that:

@@

_Acknowledgments_

The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046

>>>>

---
title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
# params:
# lang: en-GB
# author: 
# - name: Juan Carlos Castillo
#   affiliation: Department of Sociology, Universidad de Chile
#   email: juancastillov@uchile.cl
#   number: 1, 3 
# - name: Julio Iturra
#   affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
#   number: 2
# - name: Kevin Carrasco
#   affiliation: Center for Social Conflict and Cohesion Studies
#   number: 3
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. To what extent are such educational disparities justified in society? And, what are the main factors driving this kind of justification? This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy. The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.

@@

   **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
output:
  bookdown::pdf_document2:
    template: null
    number_sections: true    
    toc: false
    keep_tex: false
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion    
  bookdown::word_document2:
    number_sections: false
    # reference_docx: input/template.docx
    toc: false
  bookdown::html_document2:
    number_sections: false
linkcolor: blue
bibliography:
  - input/bib/merit-edjust.bib
csl: input/bib/apa7-singlespace.csl
editor_options:
  chunk_output_type: console
geometry: margin=1in
header-includes:
  - \usepackage{times}           # Times New Roman
  - \hypersetup {colorlinks = true, linkcolor = blue, urlcolor = blue}    
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
always_allow_html: true # for word output
fontsize: 12pt
papersize: a4
---

@@

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

@@

The Chilean case is particularly relevant when it comes to the study of inequalities in education. The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system [@bellei_estudio_2013; @corvalanMercadoEscolarLibertad2016]. The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation [@joiko_cuasimercado_2019]. Previous research in Chile has emphasized the concept of an "educational market" [@corvalanMercadoEscolarLibertad2016], characterized by selection, competition, and the provision of goods and services associated with payment capacities. Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences [@ramos_educacion_2022]. Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.

@@

Our second argument is based on the evidence that relates rational interests with justifications and preferences [@sonhing_failure_2019], expecting that:

@@

<!-- _Acknowledgments_ -->

<!-- The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046 -->
